---
title: "NIRApost Tutorial"
author: "Wang Fei, Wu Yiming, Wu Yibo, Zhu Tingshao"
date: "2026-08-30"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{NIRApost Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

NIRApost is an R package designed for post-processing of the NodeIdentifyR
Algorithm (NIRA) applied to Ising network models in psychological research.
It provides **two complementary workflows**:

**Workflow 1 — Monte Carlo simulation (Steps 1–5).** The classical NIRA
pipeline: network estimation, simulation of original and post-intervention
samples, permutation testing, and repeated-simulation stability assessment.
Its post-processing validation comprises three checks:

1. **Moderation effect testing** (`runMgmmAnalysis`) — a theoretical
   prerequisite that must be completed *before* running NIRA
2. **Permutation testing** (`permutationNIRAtest`, `plotMeanNIRA`) — formal
   significance testing with multiple-comparison correction
3. **Stability testing** (`stabilityNIRAtest`, `findMaxN`, `plotMaxN`) —
   verification that results are reproducible across independent simulation
   runs

**Workflow 2 — Analytical solution (new in v1.2.0).** A single function,
`analyticalNIRAtest()`, computes the mean number of activated nodes *exactly*
by enumerating all 2^p Ising microstates, thereby eliminating Monte Carlo
error. Because the analytical result is exact given the network parameters,
the only remaining uncertainty is **sampling error**, which is quantified by
resampling the original dataset — bootstrap-based inference is returned in
`$stat` and a random-target test in `$random`. This workflow is especially
suited to networks of small or moderate size (p ≲ 20).

Both workflows share the moderation effect test (`runMgmmAnalysis`) as a
prerequisite.

The package is used in conjunction with **nodeIdentifyR**. For full details,
please refer to the companion tutorial article:

> *Simulation Intervention for Cross-Sectional Network Models: An R Tutorial*

---

## Package Load

```{r package_load}
library(NIRApost)       # Load NIRApost Package
library(bootnet)        # Load bootnet Package
library(nodeIdentifyR)  # Load nodeIdentifyR Package
library(dplyr)          # Load dplyr Package
```

---

## Workflow 1: Monte Carlo Simulation

The following steps (1–5) implement the classical Monte Carlo NIRA pipeline.
Step 1 (network estimation) is shared by both workflows.

---

## Step 1: Import Data and Fit Ising Network

We use the built-in `single_gds` dataset (GAD-7 anxiety scale, N = 2,404).
An Ising network is estimated via `bootnet::estimateNetwork()` with the "AND"
rule. Two components are extracted:

- **Edge weight matrix** — pairwise connection weights between all nodes
- **Threshold parameter vector** — intercept terms from each node's logistic
  regression in the Ising model

```{r Ising_Network_Fit}
# Load the built-in GAD-7 anxiety dataset
data("single_gds")

# Fit the Ising model
gs_fit <- estimateNetwork(single_gds,
                           default = "IsingFit",
                           rule    = "AND")

# Extract edge weight matrix and threshold parameters
edgeWeightMatrix <- gs_fit$graph
thresholdVector  <- gs_fit$intercepts
```

---

## Step 2: Moderation Effect Test (Prerequisite for NIRA)

### Why This Must Be Done First

NIRA perturbs one node's threshold while keeping all edge weights unchanged.
This is valid **only if** no node moderates other pairwise edges. If node A
moderates the B–C relationship, then changing A's threshold implicitly alters
the B–C edge weight, making the original edge weight matrix theoretically
inappropriate for post-intervention simulation.

**This test must be run before NIRA.** An empty `significant_moderators` list
confirms the assumption holds.

### Implementation

`runMgmmAnalysis()` iteratively treats each node as a candidate moderator,
fits a Moderated Mixed Graphical Model (MGMM) via `mgm::mgm()`, and uses
resampling via `mgm::resample()` to retain only stable effects (95% CI
excluding zero).

| Parameter | Description |
|---|---|
| `data` | Dataset as a numeric matrix |
| `plotResults` | `TRUE` exports per-moderator network plots |
| `rule` | `"AND"` (stringent, default) or `"OR"` (inclusive) |
| `lambdaGam` | EBIC regularization parameter (default: 0.25) |
| `nB` | Resampling iterations (use ≥ 100 for publication) |

```{r Moderation_Effect_Test}
# nB = 10 for demonstration only; use >= 100 for publication
sig_moder <- runMgmmAnalysis(
  data        = as.matrix(single_gds),
  plotResults = FALSE,
  rule        = "AND",
  lambdaGam   = 0.25,
  nB          = 10,
  ncores                    = 6
)

# Empty list => no moderation effects => NIRA is appropriate
print(sig_moder$significant_moderators)
```

---

## Step 3: NIRA Simulation Intervention

With the moderation assumption confirmed, we proceed with NIRA using
functions from **nodeIdentifyR**.

`simulateResponses()` generates N + 1 datasets: for each of the N nodes,
one dataset with that node's threshold reduced by 2 SDs (alleviating
intervention), plus one unperturbed original dataset.

To perform aggravating interventions, change `"alleviating"` to
`"aggravating"` throughout.

```{r Simulated_alleviate_intervention}
set.seed(2025)

# Generate N+1 simulated samples
gs_IsingSamples <- simulateResponses(
  edge_weight                = edgeWeightMatrix,
  thresholds                 = thresholdVector,
  perturbation_type          = "alleviating",
  amount_of_SDs_perturbation = 2
)

# Calculate sum scores for each participant in each simulated sample
gs_sumIsingSamples <- calculateSumScores(gs_IsingSamples)

# Convert to long format for plotting
gs_sumIsingSamplesLong <- prepareDFforPlottingAndANOVA(gs_sumIsingSamples)

# Plot mean total scores (ascending order for alleviating interventions)
plotSumScores(
  sum_scores_long   = gs_sumIsingSamplesLong,
  perturbation_type = "alleviating"
)
```

---

## Step 4: Permutation Test

### Why Visual Inspection Is Insufficient

The `plotSumScores` output has four critical limitations:

1. **No formal evidence** — visual comparison cannot confirm statistical
   significance
2. **CI overlap is unreliable** — checking 95% error bar overlap is an
   inaccurate significance criterion
3. **Magnitude sensitivity** — smaller perturbations make visual assessment
   highly unreliable
4. **Multiple-comparison inflation** — N simultaneous comparisons with no
   correction inflate the Type I error rate

### Implementation

`permutationNIRAtest()` calls `permutationTest()` for every node in batch,
applies multiple-comparison correction, and returns a unified statistics
table and plotting-ready data frame. `plotMeanNIRA()` visualises the result
as a line plot with 95% CIs and significance stars
(*** p < 0.001, ** p < 0.01, * p < 0.05).

| Parameter | Description |
|---|---|
| `gs_sumIsingSamplesLong` | Long-format output from `prepareDFforPlottingAndANOVA()` |
| `method` | Multiple-comparison correction: `"bonferroni"` (default), `"BH"`, `"holm"`, etc. |

```{r permutationTest}
# Batch permutation tests with Bonferroni correction
stat_result <- permutationNIRAtest(gs_sumIsingSamplesLong,
                                    method = "bonferroni")

# View per-node statistics (mean, SD, SE, CI, Cohen's d, p, p.adjust)
head(stat_result$stat)

# Visualise means with 95% CIs and significance stars
plotMeanNIRA(stat_result, "alleviating")
```

---

## Step 5: Stability Test

### Why Stability Testing Is Necessary

A single simulation run may give different results depending on the random
seed. Only results consistent across many independent simulation runs can be
trusted.

**Note**: this procedure performs *repeated simulation* — independently
re-running the full simulation from scratch using the original network
parameters. The observed data are never resampled.

### Implementation

`stabilityNIRAtest()` repeats the full simulation `nReps` times independently.
`findMaxN()` aggregates how often each node ranked in the top-n positions
across all repetitions. `plotMaxN()` visualises the stability as a stacked
bar plot.

| Parameter | Description |
|---|---|
| `edge_weights` | Edge weight matrix from the Ising network |
| `thresholds` | Threshold parameter vector |
| `perturbation_type` | `"aggravating"` or `"alleviating"` |
| `amount_of_SDs_perturbation` | Perturbation magnitude in SDs |
| `nReps` | Number of independent simulation repetitions |
| `parallel` | `TRUE` enables parallel execution (recommended for nReps ≥ 100) |
| `ncores` | CPU cores (`NULL` = auto-detect) |

```{r stability_test}
# Parallel execution (recommended)
# nReps = 100 for demonstration; use >= 1000 for publication
simResult <- stabilityNIRAtest(
  edge_weights               = edgeWeightMatrix,
  thresholds                 = thresholdVector,
  perturbation_type          = "alleviating",
  amount_of_SDs_perturbation = 2,
  nReps                      = 100,
  parallel                   = TRUE,
  ncores                     = 6
)

# Serial execution (uncomment to use; displays a progress bar)
# simResult <- stabilityNIRAtest(
#   edgeWeightMatrix, thresholdVector,
#   "alleviating", 2,
#   nReps    = 100,
#   parallel = FALSE,
#   ncores   = NULL
# )

# Compute rank frequencies across all simulation repetitions
repeat_node <- findMaxN(simResult, n = 7)
print(repeat_node)

# Stacked bar plot of node ranking stability
plotMaxN(repeat_node, low = 1, high = 7)
```

---

## Workflow 2: Analytical Solution — `analyticalNIRAtest`

### Why an Analytical Approach?

Cui et al. (2026) showed that the expected number of activated nodes in an
Ising network can be obtained *exactly* by enumerating all 2^p microstates,
without Monte Carlo simulation. Each activation state a has a Boltzmann
weight given by its Hamiltonian, and the mean number of activated nodes is
the Boltzmann-weighted average over all 2^p states. Because the result is
exact, **Monte Carlo error is eliminated**, and the computation is extremely
fast for small to moderate networks.

The only remaining source of uncertainty is **sampling error**: the network
parameters (W, m) are estimated from a finite sample. `analyticalNIRAtest()`
quantifies this uncertainty by resampling the **original dataset** — each
resample re-estimates the network and recomputes the exact solution — and
returns bootstrap-based inference and a random-target test in a single call.

### The `analyticalNIRAtest()` Function

| Parameter | Description |
|---|---|
| `data` | Binary (0/1) data; rows = participants, columns = nodes |
| `perturbation_type` | `"aggravating"` (threshold up) or `"alleviating"` (threshold down) |
| `edge_weights`, `thresholds` | Optional pre-estimated network parameters (auto-estimated if `NULL`) |
| `amount_of_SDs_perturbation` | Perturbation magnitude in SDs of thresholds (default: 2) |
| `nResample` | Number of resamples of the original data (default: 5000) |
| `method` | Multiple-comparison correction for the bootstrap p-values (default: `"bonferroni"`) |
| `seed` | Random seed |
| `parallel`, `ncores` | Parallel computation settings |

The function returns a list with four main elements:

- **`$stat`** — bootstrap-based inference per node: exact mean activation,
  intervention effect (Δ), Cohen's *d*, standard error, 95% confidence
  intervals, *p*-value for H0: Δ = 0, adjusted *p*-value, and `percenttop_1`
  (the proportion of resamples in which each node ranked first — an
  analytical stability metric).
- **`$random`** — a random-target test: the (uncorrected) probability that a
  randomly chosen node has an effect at least as large as node *i*'s effect,
  answering "is node *i* a better intervention target than a random node?"
- **`$exact`** — the exact mean activation for the original network and each
  one-node intervention.
- **`$plot`** — a ggplot of the exact mean activation with 95% CI error bars
  and significance stars.

```{r analytical_workflow}
# Analytical NIRA: exact mean activation + resampling-based uncertainty
res <- analyticalNIRAtest(
  data                      = single_gds,
  perturbation_type         = "alleviating",
  edge_weights              = edgeWeightMatrix,
  thresholds                = thresholdVector,
  amount_of_SDs_perturbation = 2,
  nResample                 = 1000,   # 5000 recommended for publication
  method                    = "bonferroni",
  seed                      = 2025,
  parallel                  = TRUE,
  ncores                    = 6
)

# Bootstrap-based inference: effect, SE, CI, p, percenttop_1
head(res$stat)

# Random-target test (uncorrected): is each node better than a random target?
res$random

# Exact mean activation for the original network and each intervention
res$exact

# Plot of exact mean activation with 95% CI and significance stars
res$plot
```

### How to Choose Between the Two Workflows?

| Situation | Recommended workflow |
|---|---|
| Small or moderate network (p ≲ 20); exact, fast results desired | **Workflow 2** (`analyticalNIRAtest`) |
| Large network where 2^p enumeration is infeasible | **Workflow 1** (Monte Carlo) |
| Full simulation-based permutation / stability infrastructure needed | **Workflow 1** |

---

## Workflow 3: Bridge (Two-Community) NIRA — `analyticalBridgeNIRAtest`

### What Is a Bridge Intervention?

In bridge-network research the network is partitioned into **two communities**
(e.g., health-behaviour nodes vs. mental-health symptom nodes), and the
question is *which node of one community, when intervened upon, most strongly
changes the activation of the other community*. The `Bridge_network_NIRA.R`
script computes networktools bridge centrality first, then uses NIRA to
intervene on each node of the source community (JKXW) and measures the total
score of the outcome community (MH) as the downstream outcome.

`analyticalBridgeNIRAtest()` is the exact analytical version of this
procedure: it enumerates all 2^p microstates once, but the outcome statistic
is restricted to the **outcome community's** activation count
(Σ_{j∈B} s_j + p_B)/2. Intervening on a source node reweights the microstates
exactly, so no re-enumeration is needed for the p_A source interventions.

For this workflow we use a different dataset: `bridge_network.xlsx`
(health-behaviour and mental-health items) is read and cleaned into
`GS_non_null` as in `Bridge_network_NIRA.R`, and its Ising network is fitted
with `bootnet::estimateNetwork()` to obtain `edgeWeightMatrix` and
`thresholdVector`.

### Prerequisite: Moderation-Effect Test on the Whole Network

Before any bridge intervention, the moderation-effect test
(`runMgmmAnalysis()`) must be run on the **entire** network (both communities
together), exactly as in Workflow 1. The threshold-shift intervention model
is a property of the full joint distribution, and an intervention on a source
node propagates through the whole network — so a moderator *anywhere* would
change edge weights during the intervention and invalidate the model. An
empty `significant_moderators` list confirms the assumption holds.

```{r bridge_moderation}
# Whole-network moderation test (Step 3 of Bridge_network_NIRA.R).
# ncores > 1 parallelises the p candidate moderator nodes (identical results
# to the serial run); set ncores = 1 for serial execution.
sig_moder <- runMgmmAnalysis(
  data        = as.matrix(GS_non_null),
  plotResults = FALSE,
  rule        = "AND",
  lambdaGam   = 0.25,
  nB          = 10,          # use >= 100 for publication
  ncores      = 6            # parallel over moderator nodes
)
# Empty list => no significant moderation => bridge NIRA is appropriate
print(sig_moder$significant_moderators)
```

```{r bridge_workflow}
groups <- rep(c("JKXW", "MH"), c(8, 3))      # 8 source + 3 outcome nodes

res2 <- analyticalBridgeNIRAtest(
  data                      = GS_non_null,    # binary data (N x p)
  groups                    = groups,
  intervention_group        = "JKXW",         # intervene on source nodes
  outcome_group             = "MH",           # measure outcome community
  perturbation_type         = "aggravating",
  edge_weights              = edgeWeightMatrix,  # from bootnet (fit$graph)
  thresholds                = thresholdVector,   # from bootnet (fit$intercepts)
  amount_of_SDs_perturbation = 2,
  nResample                 = 1000,           # 5000 recommended for publication
  method                    = "bonferroni",
  seed                      = 2025,
  parallel                  = TRUE,
  ncores                    = 6
)
```

### How to Read the Output

**`$exact`** — exact mean number of activated *outcome* nodes under the
original network and under each one-node source intervention:

```{r bridge_exact}
res2$exact
```

**`$stat`** — per-source-node inference on the outcome community: exact mean
activation, 95% CI, intervention effect Δ, Cohen's *d*, bootstrap SE/CI, the
bootstrap *p*-value for H0: Δ = 0, its adjusted value, and `percenttop_1`
(the proportion of resamples in which the node had the largest absolute
effect on the outcome community):

```{r bridge_stat}
res2$stat
```

**`$random`** — the random-target test, restricted to the *source* community:
for each source node, the (uncorrected) probability that a randomly chosen
source node would produce an effect at least as large.

```{r bridge_random}
res2$random
```

**`$plot`** — the exact outcome-community mean activation of each condition
with 95% CI error bars and significance stars.

### Interpretation: What to Look At

- **`mean_activation` is on the outcome-community scale** (0–p_B), not the
  full-network scale. Read it as "expected number of activated *outcome*
  nodes", not "total activated nodes". In the example, the MH community
  (depression, anxiety, stress) is active at only ~0.65 nodes on average,
  so most participants report at most one of the three symptoms.
- **`effect` / Δ** is the difference between the outcome-community mean after
  the intervention and the original value (from `$exact`). Values near 0 mean
  the source node barely propagates to the outcome community (weak or absent
  bridge edges).
- **`p.adjust` tests H0: Δ = 0.** Because the analytical solution makes every
  single-node effect deterministically non-zero, these *p*-values are usually
  trivially non-significant for every node and are *not* the discriminating
  output.
- **`percenttop_1` is the discriminating metric.** It is the analytical
  analogue of `findMaxN()`: the proportion of resamples in which the node
  ranked first (largest absolute effect on the outcome community). A node
  with `percenttop_1` far above the others is the most consistently strong
  bridge target. In the example, 睡太少 (too little sleep) ranked first in
  51.4% of resamples, ahead of 不吃早餐 (17.4%) and 独居 (12.9%).
- **`$random`** answers "is node *i* a better target than a randomly chosen
  source node?" within the source community. Large (uncorrected) *p*-values
  mean no source node is demonstrably superior to a random source target.

### How to Report in a Paper

A concise bridge-NIRA results paragraph could read:

> Using the analytical bridge-NIRA workflow, we intervened on each
> health-behaviour node and computed the exact expected number of activated
> mental-health symptoms (depression, anxiety, stress). In the original
> network the outcome community was active at a mean of 0.65 nodes.
> Aggravating intervention on 7 of the 8 behaviour nodes produced negligible
> changes in mental-health activation (Δ < 0.004, all Bonferroni-adjusted
> *p* = 1). Only intervening on too-little-sleep increased the expected
> number of activated symptoms, by Δ = 0.039 (Cohen's *d* = 0.041), but this
> effect was not statistically reliable (95% CI [−0.001, 0.158]; bootstrap
> *p* = 0.489, adjusted *p* = 1). Too-little-sleep nevertheless ranked first
> in 51.4% of the 1000 resamples (`percenttop_1`), suggesting it is the most
> consistently promising bridge target, although its superiority over a
> randomly chosen behaviour node did not reach significance.

**Reporting checklist.** (1) State the outcome community and the scale of
`mean_activation` (0–p_B). (2) Report the original pre-intervention value.
(3) Report Δ, Cohen's *d*, 95% CI and adjusted *p*-value for each target (or
at least the top-ranked ones). (4) Interpret `percenttop_1` as ranking
stability, not significance. (5) Note the model assumption that the
moderation-effect test (`runMgmmAnalysis()`) found no moderators, so a
single-node intervention leaves all edge weights unchanged.

---

## Summary: Three Workflows

| Workflow | Key steps | Primary function(s) | Purpose |
|---|---|---|---|
| **Workflow 1 (Monte Carlo)** | Moderation → Simulation → Permutation → Stability | `runMgmmAnalysis()`, `permutationNIRAtest()`, `stabilityNIRAtest()`, `findMaxN()`, `plotMaxN()` | Quantify Monte Carlo error |
| **Workflow 2 (Analytical)** | Moderation → Analytical NIRA | `runMgmmAnalysis()`, `analyticalNIRAtest()` | Exact results + sampling-error uncertainty |
| **Workflow 3 (Bridge)** | Moderation → Analytical bridge NIRA | `runMgmmAnalysis()`, `analyticalBridgeNIRAtest()` | Exact cross-community (source → outcome) intervention effects |

---

## Session Info

```{r session_info, eval=TRUE, echo=FALSE}
sessionInfo()
```
