---
title: "NIRApost Tutorial"
author: "Wang Fei, Wu Yiming, Wu Yibo, Zhu Tingshao"
date: "2026-08-30"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{NIRApost Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

NIRApost is an R package designed for post-processing of the NodeIdentifyR
Algorithm (NIRA) applied to Ising network models in psychological research.
It provides **two complementary workflows**:

**Workflow 1 — Monte Carlo simulation (Steps 1–5).** The classical NIRA
pipeline: network estimation, simulation of original and post-intervention
samples, permutation testing, and repeated-simulation stability assessment.
Its post-processing validation comprises three checks:

1. **Moderation effect testing** (`runMgmmAnalysis`) — a theoretical
   prerequisite that must be completed *before* running NIRA
2. **Permutation testing** (`permutationNIRAtest`, `plotMeanNIRA`) — formal
   significance testing with multiple-comparison correction
3. **Stability testing** (`stabilityNIRAtest`, `findMaxN`, `plotMaxN`) —
   verification that results are reproducible across independent simulation
   runs

**Workflow 2 — Analytical solution (new in v1.2.0).** A single function,
`analyticalNIRAtest()`, computes the mean number of activated nodes *exactly*
by enumerating all 2^p Ising microstates, thereby eliminating Monte Carlo
error. Because the analytical result is exact given the network parameters,
the only remaining uncertainty is **sampling error**, which is quantified by
resampling the original dataset — bootstrap-based inference is returned in
`$stat` and a random-target test in `$random`. This workflow is especially
suited to networks of small or moderate size (p ≲ 20).

Both workflows share the moderation effect test (`runMgmmAnalysis`) as a
prerequisite.

The package is used in conjunction with **nodeIdentifyR**. For full details,
please refer to the companion tutorial article:

> *Simulation Intervention for Cross-Sectional Network Models: An R Tutorial*

---

## Package Load

```{r package_load}
library(NIRApost)       # Load NIRApost Package
library(bootnet)        # Load bootnet Package
library(nodeIdentifyR)  # Load nodeIdentifyR Package
library(dplyr)          # Load dplyr Package
```

---

## Workflow 1: Monte Carlo Simulation

The following steps (1–5) implement the classical Monte Carlo NIRA pipeline.
Step 1 (network estimation) is shared by both workflows.

---

## Step 1: Import Data and Fit Ising Network

We use the built-in `single_gds` dataset (GAD-7 anxiety scale, N = 2,404).
An Ising network is estimated via `bootnet::estimateNetwork()` with the "AND"
rule. Two components are extracted:

- **Edge weight matrix** — pairwise connection weights between all nodes
- **Threshold parameter vector** — intercept terms from each node's logistic
  regression in the Ising model

```{r Ising_Network_Fit}
# Load the built-in GAD-7 anxiety dataset
data("single_gds")

# Fit the Ising model
gs_fit <- estimateNetwork(single_gds,
                           default = "IsingFit",
                           rule    = "AND")

# Extract edge weight matrix and threshold parameters
edgeWeightMatrix <- gs_fit$graph
thresholdVector  <- gs_fit$intercepts
```

---

## Step 2: Moderation Effect Test (Prerequisite for NIRA)

### Why This Must Be Done First

NIRA perturbs one node's threshold while keeping all edge weights unchanged.
This is valid **only if** no node moderates other pairwise edges. If node A
moderates the B–C relationship, then changing A's threshold implicitly alters
the B–C edge weight, making the original edge weight matrix theoretically
inappropriate for post-intervention simulation.

**This test must be run before NIRA.** An empty `significant_moderators` list
confirms the assumption holds.

### Implementation

`runMgmmAnalysis()` iteratively treats each node as a candidate moderator,
fits a Moderated Mixed Graphical Model (MGMM) via `mgm::mgm()`, and uses
resampling via `mgm::resample()` to retain only stable effects (95% CI
excluding zero).

| Parameter | Description |
|---|---|
| `data` | Dataset as a numeric matrix |
| `plotResults` | `TRUE` exports per-moderator network plots |
| `rule` | `"AND"` (stringent, default) or `"OR"` (inclusive) |
| `lambdaGam` | EBIC regularization parameter (default: 0.25) |
| `nB` | Resampling iterations (use ≥ 100 for publication) |

```{r Moderation_Effect_Test}
# nB = 10 for demonstration only; use >= 100 for publication
sig_moder <- runMgmmAnalysis(
  data        = as.matrix(single_gds),
  plotResults = FALSE,
  rule        = "AND",
  lambdaGam   = 0.25,
  nB          = 10
)

# Empty list => no moderation effects => NIRA is appropriate
print(sig_moder$significant_moderators)
```

---

## Step 3: NIRA Simulation Intervention

With the moderation assumption confirmed, we proceed with NIRA using
functions from **nodeIdentifyR**.

`simulateResponses()` generates N + 1 datasets: for each of the N nodes,
one dataset with that node's threshold reduced by 2 SDs (alleviating
intervention), plus one unperturbed original dataset.

To perform aggravating interventions, change `"alleviating"` to
`"aggravating"` throughout.

```{r Simulated_alleviate_intervention}
set.seed(2025)

# Generate N+1 simulated samples
gs_IsingSamples <- simulateResponses(
  edge_weight                = edgeWeightMatrix,
  thresholds                 = thresholdVector,
  perturbation_type          = "alleviating",
  amount_of_SDs_perturbation = 2
)

# Calculate sum scores for each participant in each simulated sample
gs_sumIsingSamples <- calculateSumScores(gs_IsingSamples)

# Convert to long format for plotting
gs_sumIsingSamplesLong <- prepareDFforPlottingAndANOVA(gs_sumIsingSamples)

# Plot mean total scores (ascending order for alleviating interventions)
plotSumScores(
  sum_scores_long   = gs_sumIsingSamplesLong,
  perturbation_type = "alleviating"
)
```

---

## Step 4: Permutation Test

### Why Visual Inspection Is Insufficient

The `plotSumScores` output has four critical limitations:

1. **No formal evidence** — visual comparison cannot confirm statistical
   significance
2. **CI overlap is unreliable** — checking 95% error bar overlap is an
   inaccurate significance criterion
3. **Magnitude sensitivity** — smaller perturbations make visual assessment
   highly unreliable
4. **Multiple-comparison inflation** — N simultaneous comparisons with no
   correction inflate the Type I error rate

### Implementation

`permutationNIRAtest()` calls `permutationTest()` for every node in batch,
applies multiple-comparison correction, and returns a unified statistics
table and plotting-ready data frame. `plotMeanNIRA()` visualises the result
as a line plot with 95% CIs and significance stars
(*** p < 0.001, ** p < 0.01, * p < 0.05).

| Parameter | Description |
|---|---|
| `gs_sumIsingSamplesLong` | Long-format output from `prepareDFforPlottingAndANOVA()` |
| `method` | Multiple-comparison correction: `"bonferroni"` (default), `"BH"`, `"holm"`, etc. |

```{r permutationTest}
# Batch permutation tests with Bonferroni correction
stat_result <- permutationNIRAtest(gs_sumIsingSamplesLong,
                                    method = "bonferroni")

# View per-node statistics (mean, SD, SE, CI, Cohen's d, p, p.adjust)
head(stat_result$stat)

# Visualise means with 95% CIs and significance stars
plotMeanNIRA(stat_result, "alleviating")
```

---

## Step 5: Stability Test

### Why Stability Testing Is Necessary

A single simulation run may give different results depending on the random
seed. Only results consistent across many independent simulation runs can be
trusted.

**Note**: this procedure performs *repeated simulation* — independently
re-running the full simulation from scratch using the original network
parameters. The observed data are never resampled.

### Implementation

`stabilityNIRAtest()` repeats the full simulation `nReps` times independently.
`findMaxN()` aggregates how often each node ranked in the top-n positions
across all repetitions. `plotMaxN()` visualises the stability as a stacked
bar plot.

| Parameter | Description |
|---|---|
| `edge_weights` | Edge weight matrix from the Ising network |
| `thresholds` | Threshold parameter vector |
| `perturbation_type` | `"aggravating"` or `"alleviating"` |
| `amount_of_SDs_perturbation` | Perturbation magnitude in SDs |
| `nReps` | Number of independent simulation repetitions |
| `parallel` | `TRUE` enables parallel execution (recommended for nReps ≥ 100) |
| `ncores` | CPU cores (`NULL` = auto-detect) |

```{r stability_test}
# Parallel execution (recommended)
# nReps = 100 for demonstration; use >= 1000 for publication
simResult <- stabilityNIRAtest(
  edge_weights               = edgeWeightMatrix,
  thresholds                 = thresholdVector,
  perturbation_type          = "alleviating",
  amount_of_SDs_perturbation = 2,
  nReps                      = 100,
  parallel                   = TRUE,
  ncores                     = 6
)

# Serial execution (uncomment to use; displays a progress bar)
# simResult <- stabilityNIRAtest(
#   edgeWeightMatrix, thresholdVector,
#   "alleviating", 2,
#   nReps    = 100,
#   parallel = FALSE,
#   ncores   = NULL
# )

# Compute rank frequencies across all simulation repetitions
repeat_node <- findMaxN(simResult, n = 7)
print(repeat_node)

# Stacked bar plot of node ranking stability
plotMaxN(repeat_node, low = 1, high = 7)
```

---

## Workflow 2: Analytical Solution — `analyticalNIRAtest`

### Why an Analytical Approach?

Cui et al. (2026) showed that the expected number of activated nodes in an
Ising network can be obtained *exactly* by enumerating all 2^p microstates,
without Monte Carlo simulation. Each activation state a has a Boltzmann
weight given by its Hamiltonian, and the mean number of activated nodes is
the Boltzmann-weighted average over all 2^p states. Because the result is
exact, **Monte Carlo error is eliminated**, and the computation is extremely
fast for small to moderate networks.

The only remaining source of uncertainty is **sampling error**: the network
parameters (W, m) are estimated from a finite sample. `analyticalNIRAtest()`
quantifies this uncertainty by resampling the **original dataset** — each
resample re-estimates the network and recomputes the exact solution — and
returns bootstrap-based inference and a random-target test in a single call.

### The `analyticalNIRAtest()` Function

| Parameter | Description |
|---|---|
| `data` | Binary (0/1) data; rows = participants, columns = nodes |
| `perturbation_type` | `"aggravating"` (threshold up) or `"alleviating"` (threshold down) |
| `edge_weights`, `thresholds` | Optional pre-estimated network parameters (auto-estimated if `NULL`) |
| `amount_of_SDs_perturbation` | Perturbation magnitude in SDs of thresholds (default: 2) |
| `nResample` | Number of resamples of the original data (default: 5000) |
| `method` | Multiple-comparison correction for the bootstrap p-values (default: `"bonferroni"`) |
| `seed` | Random seed |
| `parallel`, `ncores` | Parallel computation settings |

The function returns a list with four main elements:

- **`$stat`** — bootstrap-based inference per node: exact mean activation,
  intervention effect (Δ), Cohen's *d*, standard error, 95% confidence
  intervals, *p*-value for H0: Δ = 0, adjusted *p*-value, and `percenttop_1`
  (the proportion of resamples in which each node ranked first — an
  analytical stability metric).
- **`$random`** — a random-target test: the (uncorrected) probability that a
  randomly chosen node has an effect at least as large as node *i*'s effect,
  answering "is node *i* a better intervention target than a random node?"
- **`$exact`** — the exact mean activation for the original network and each
  one-node intervention.
- **`$plot`** — a ggplot of the exact mean activation with 95% CI error bars
  and significance stars.

```{r analytical_workflow}
# Analytical NIRA: exact mean activation + resampling-based uncertainty
res <- analyticalNIRAtest(
  data                      = single_gds,
  perturbation_type         = "alleviating",
  edge_weights              = edgeWeightMatrix,
  thresholds                = thresholdVector,
  amount_of_SDs_perturbation = 2,
  nResample                 = 1000,   # 5000 recommended for publication
  method                    = "bonferroni",
  seed                      = 2025,
  parallel                  = TRUE,
  ncores                    = 6
)

# Bootstrap-based inference: effect, SE, CI, p, percenttop_1
head(res$stat)

# Random-target test (uncorrected): is each node better than a random target?
res$random

# Exact mean activation for the original network and each intervention
res$exact

# Plot of exact mean activation with 95% CI and significance stars
res$plot
```

### How to Choose Between the Two Workflows?

| Situation | Recommended workflow |
|---|---|
| Small or moderate network (p ≲ 20); exact, fast results desired | **Workflow 2** (`analyticalNIRAtest`) |
| Large network where 2^p enumeration is infeasible | **Workflow 1** (Monte Carlo) |
| Full simulation-based permutation / stability infrastructure needed | **Workflow 1** |

---

## Summary: Two Workflows

| Workflow | Key steps | Primary function(s) | Purpose |
|---|---|---|---|
| **Workflow 1 (Monte Carlo)** | Moderation → Simulation → Permutation → Stability | `runMgmmAnalysis()`, `permutationNIRAtest()`, `stabilityNIRAtest()`, `findMaxN()`, `plotMaxN()` | Quantify Monte Carlo error |
| **Workflow 2 (Analytical)** | Moderation → Analytical NIRA | `runMgmmAnalysis()`, `analyticalNIRAtest()` | Exact results + sampling-error uncertainty |

---

## Session Info

```{r session_info, eval=TRUE, echo=FALSE}
sessionInfo()
```
